### 2022-11-25 garbage_collect/main.tex
add garbage_collect+podc_papaer

\documentclass[11pt,letterpaper]{article}
\renewcommand{\baselinestretch}{1.8}

\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{showkeys}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{relsize}
\usepackage{enumitem}
\usepackage{bold-extra}
\renewcommand*\contentsname{Table of Contents}

@@

\renewcommand\thealgorithm{}
\newcommand{\setalglineno}[1]{%
  \setcounter{ALC@line}{\numexpr#1-1}}

@@

\begin{center}
\pagenumbering{roman}
\huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}

\vspace{2em}
\Large{\bf{Hossein Naderibeni}}

\vspace{3em}
\large A Thesis submitted to the Faculty of Graduate Studies \\ in Partial Fulfillment of the Requirements\\for the Degree of Master of Science

\vspace{2em}
Graduate Program in Computer Science,\\
York University\\
Toronto, Ontario

\vspace{2em}
November, 2022

\vspace{3em}
\textcopyright \; Hossein Naderibeni, 2022

\end{center}
\thispagestyle{empty}
\pagebreak

\normalsize
\addcontentsline{toc}{subsection}{Abstract}
\section*{Abstract}

In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem. We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.

\pagebreak

\addcontentsline{toc}{subsection}{Acknowledgements}
\section*{Acknowledgements}
\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert. Without his guidance and support, this thesis probably would not have been completed. He helped me in every step of this thesis, taught me shared data structures, helped me find the problem, design and improve the algorithm, and write it with its complications because of lots of new notions in it. Thank you again for always pushing to make it better.

I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments but also for his academic advice and motivations.

Besides my supervisors, I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.

Last but not least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study. I want to thank my brother for being there for me whenever I needed help.

I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.

I dedicate this paper to the people in Iran and all the people who are fighting for their freedom.

\pagebreak
\addcontentsline{toc}{subsection}{Table of Contents}
\tableofcontents

\pagebreak
\addcontentsline{toc}{subsection}{List of Tables}
\listoftables

\pagebreak
\addcontentsline{toc}{subsection}{List of Figures}
\listoffigures
\newpage

\pagebreak

\pagenumbering{arabic}
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same. In this run, $P$'s update is overwritten, which makes the queue inconsistent. One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored. \texttt{Dequeue} removes and returns the first element in the queue. If the queue is empty, it returns \texttt{null}. In a concurrent version of the queue data structure, operations do not happen one at a time. The question that may arise is, ``What properties matter for the implementation of a shared data structure?'' Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy based on how powerful objects are to solve the consensus problem among processes. Objects like LL/SC or CAS can be used to reach a consensus among any number of processors. A CAS object provides an atomic Compare\&Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}; otherwise it returns \nf{false}.  In this work, we use Compare\&Swap (\nf{CAS}) objects to synchronize among processes.

% \bf{Concurrent execution: I didn't understed interleaving steps of processes}

The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response. Figure \ref{fig::goodexample} depicts an example of an execution on a linearizable  queue that is initially empty. The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation. Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}. The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/good}
  \caption[An example of a linearizable execution.]{\label{fig::goodexample}An example of a linearizable execution. Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
\end{figure}

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption[An example of a non-linearizable execution.]{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable. \texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before \texttt{Enqueue(A)} and return \nf{null} or can occur after \texttt{Enqueue(A)} and return \texttt{A}.}
\end{figure}

There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom. An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm, there might be an operation that takes infinitely many steps but never terminates.

We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps. We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized. Processes can do queries to get information about the linearization stored in the root. We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized. \nf{CAS} instructions cost more than other instructions for the processor. Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case. We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations. Furthermore, we do not address garbage collection but we describe a way we think it can be handled.

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation. Previous lock-free queues and their common problem are presented in Section 2.1. In Section 2.2, we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.

In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. We also discuss the motivation and requirements in our design to achieve poly-log time. Section 3.2 contains the algorithm itself and there is an example of a concurrent execution of the algorithm described in Section 3.3.

We prove the correctness of our queue in Chapter 4 by showing it is linearizable.

Chapter 5 is devoted to a complexity analysis for our implementation. We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue. In the end, we prove our queue is wait-free.

Finally, we give some concluding remarks in Chapter 6. It contains how we can improve our algorithm's memory usage and how to use its idea in designing other wait-free data structures.

\pagebreak
\section{Related Work}
In this section, we look at previous lock-free queues. The amortized step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the worst case, where all $p$ processes try to do an enqueue or a dequeue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

\subsection{List-based Queues}
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. A version of it is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}). A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue. To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the amortized number of steps $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.4]{pics/msqueue}
  \caption[MS-queue structure.]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. In the first diagram, the first element has been dequeued. Red arrows show new pointers and gray dashed arrows show the old pointers.}
\end{figure}

Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the \it{elimination} technique. The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue. An Elimination Queue consists of an MS-queue augmented with an elimination array. Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue. Their algorithm makes it possible for long-running operations to eliminate an opposing operation. The empirical evaluation showed the throughput of their work is better than the MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.

Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue. Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket. Each basket has a time interval in which all its nodes' enqueue operations overlap. Since the operations in a basket are concurrent, we can order them in any way. Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations. However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption[Baskets queue.]{At some time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}. To order the operations in a basket, processes perform another \nf{CAS}. The successful process will be the next one in the basket, and so on.}
\end{figure}

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue. MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous node to the new node. They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue. As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one. The amortized complexity is still $\Omega(p)$ \texttt{CAS}es.

Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining. The key idea behind flat combining is to allow a combiner who has acquired the global lock on the data structure to learn about the requests of threads on the queue, combine them and apply the combined results on the data structure. Their queue is linearizable but not lock-free and they present experiments that show their algorithm performs well in some situations.

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. The queue is stored in a shared array where head and tail pointers point to the current elements in the queue. When the array is full, an empty array is linked to the array and tail pointers are updated. A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array. Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}). A process updates the position of the pointers after it does an operation. One process might go to sleep before setting the pointers, so the pointers might be behind their real places. They mention how to scan the arrays to update pointers while doing an operation. A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the amortized step (and \nf{CAS}) complexity remains $\Omega(p)$. Their design is lock-free but not wait-free.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/sundell}
    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}.]{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
\end{figure}

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom. Their amortized step complexity is $\Omega(p)$ because of the helping mechanism.

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures. In their queue, operations  return future objects instead of responses. Later when the response is needed, it can be evaluated from the future object. They also define a weaker linearizability condition such that each operation can be linearized between its invocation and when its future is evaluated. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue. They use some properties of the queue size before and after a batch, similar to a part of our work. Their queue is not wait-free: in fact, if the batch sizes are 1, then the queue is like MS-queue.

Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. The fast path  ensures good performance while the slow path ensures termination. They show that these two paths do not affect each other and the queue remains consistent. If a process makes no progress, other processes help its slow path to finish. The helping phase suffers from the CAS Retry Problem because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the amortized complexity cannot be better than $\Omega(p)$.

The CAS Retry Problem is not limited to list-based queues; array-based queues also share it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  Our motivation is to overcome this problem and present a wait-free sublinear queue.

\subsection{Restricted Queues}

David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}. Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
The author states that to reduce memory usage to be bounded, the time per operation increases linearly.

Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a poly-logarithmic multi-enqueuer multi-dequeuer queue. Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.

\subsection{Universal Constructions and Other Poly-log Time Data Structures}
A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. The first universal construction was introduced by Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$-bit words. Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. However, their algorithm cannot be used to create a queue. We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of page 13.

% TODO: page

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes. When a process wants to do a Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number of elements in the tree. There are some similarities between designing a queue and a Fetch\&Inc object. A Fetch\&Inc object can be constructed from a queue. The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues. Also, it is one of the few wait-free data structures achieving poly-logarithmic complexity.

\subsection{Attiya--Fouren Lower Bound}
Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues. Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes. Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is that since their result is about when contention is low, their lower bound does not contradict our algorithm's complexity and we manage to reach poly-log time complexity.

\pagebreak

\section{Queue Implementation} \label{DescriptQ}

In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently. We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization. We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.

There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes. The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}). Each process has a leaf in which the operations invoked by the process are stored in order. When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root. Each node of the tree keeps an ordering of operations propagated up to it. All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering. 
\begin{figure}[hbtp]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored. Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}  
\end{figure}

To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$. We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$. We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
We sketch the argument here.

\begin{figure}[hbtp]
\begin{center}
\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{Before the \nf{Refresh}.}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.29\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
$l_1,l_2,l_3,l_4,l_5$
$r_1,r_2,r_3,r_4$ ]
\end{tikzpicture}}
  \caption{After the \nf{Refresh}.}
\end{subfigure}
\hspace{6em}
\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append. Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
\end{center}
\end{figure}

\begin{figure}[hbpt]
  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
\end{figure}

We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}). This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended. After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node. So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization. This cooperative solution allows us to overcome the CAS Retry Problem.

It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
Instead we use an implicit representation of the operations propagated together. Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}). Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}s. A block stores information about the operations propagated by a \nf{Refresh}. It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure. See Figure~\ref{fig:block} for an example. A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent. We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently. Thus, there are at most $p$ operations in a block. Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.

\begin{figure}[hbtp]
\begin{center}
\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]

\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ... \\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ... \\ \hline\end{tabular}} ] ]
\end{tikzpicture}
\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right. In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}. For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together. Since their execution time intervals overlap, they can be linearized in any order.}
\end{center}
\end{figure}

\begin{figure}[hbtp]
\begin{center}
\begin{tikzpicture}
[level 1/.style={level distance=3.3cm,sibling distance=0cm},
	level 2/.style={level distance=2.5cm,sibling distance=0cm},
	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
\begin{small}

\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
\end{small}
\end{tikzpicture}
\end{center}
\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations. Blocks between two lines $||$ are propagated together to the parent. Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively. For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child. The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child. (5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
\end{figure}

So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$. However, this approach is not enough for an efficient queue.
We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization. We explain how to implement (1) and (2) in poly-logarithmic steps.

After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation. When we reach a block in a leaf, the operation is explicitly stored there. To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block. We also store pointers to determine the range of subblocks of a block to make the binary search faster. In each block, we store the prefix sum of operations from the left child and from the right child. Moreover, for each block, we store two attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of the last left and right subblock (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and \nf{end\sub{right}} indices of a block and its previous block reduce the search range size to \textsc{O}$(p)$.

To compute the rank in the root of an operation in a leaf, we need to find the superblock of the block that the operation is in. After a block is installed in a node we store the approximate index of its superblock in it to make this faster.

\begin{figure}[hbtp]
\centering
  \includegraphics[width=4in, height=1.4in]{pics/pointers}
  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
\end{figure}

In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue. In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0. We refer to such a dequeue as a \it{null dequeue}. If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue. Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation. So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.

In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.  We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations. Consider the following operations, where operations in a cell are concurrent.
\begin{table}[hbtp]
\centering
\begin{small}
\begin{tabular}{c|c|c|c}
    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
  \end{tabular}
\end{small}
\caption{An execution on a queue separated by the operations in the root blocks.}
\end{table}

\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}. We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\renewcommand{\baselinestretch}{1.8}
\usepackage[dvipsnames]{xcolor}
\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{relsize}
\usepackage{enumitem}
\usepackage{bold-extra}
\renewcommand*\contentsname{Table of Contents}

@@

\renewcommand\thealgorithm{}
\newcommand{\setalglineno}[1]{
  \setcounter{ALC@line}{\numexpr#1-1}}

@@

\section{Feilds}
\begin{algorithm}
\caption{Tree Fields Description}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
\item {\color{red}\tt{\sl{MaxbyProcess} lastDequeuedFrom} \textsf{Index of the most recent block in the root that has been deqeueued from.}}
\end{itemize}

\Statex $\diamondsuit$ \tt{\sl{Local}}
\begin{itemize}
\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
\item {\color{red}\tt{\sl{int} garbageCollectRound}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
\item {\color{red}\tt{\sl{PBRT} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
\end{itemize}
